Noncoherent least squares estimators of carrier phase and amplitude

### 2012-08-02 other/confnoncoherent/paper.tex
Added to conf

\begin{abstract}
We consider least squares estimators of carrier phase and amplitude from a noisy communications signal.  We focus on signaling constellations that have symbols evenly distributed on the complex unit circle, such as binary phase shift keying (BPSK), quaternary phase shift keying (QPSK), and $M$-PSK.  We show, under reasonably mild conditions on the distribution of the noise, that the least squares estimator of carrier phase is strongly consistent and asymptotically normaly distributed.  However, the amplitude estimator is not consistent, and we show that it converges to a positive real number that is a function of the true carrier amplitude, the noise distribution, and the size of the constellation.  The results of Monte Carlo simulations are provided and these corroborate the theoretical results.
\end{abstract}
\begin{IEEEkeywords}
BLERG
\end{IEEEkeywords}

@@

In passband communications systems the transmitted signal typically undergoes time offset (delay), phase shift and attenuation.  These effects must be compensated for at the reciever. In this paper we assume that the time offset has been previously handled, and we focus on estimating the phase shift and attenuation.  We consider signalling constellations that have symbols evenly distributed on the complex unit circle, commonly refered to as $M$-ary phase shift keying ($M$-PSK).  In this case, the transmitted symbols take the form,
\[
s_i = e^{j u_i},
\]
where $j = \sqrt{-1}$ and $u_i$ is from the set $\{0, \tfrac{2\pi}{M}, \dots, \tfrac{2\pi(M-1)}{M}\}$ and $M$ is the size of the constellation.

We assume that time offset estimation and matched filtering have already been performed at the reciever.  The recieved symbols then take the form,
\begin{equation}\label{eq:sigmod}
y_i = a_0 s_i + w_i, \qquad i = 1, \dots, L,
\end{equation}
where $w_i$ is noise and $a_0 = \rho_0 e^{j\theta_0}$ is a complex number representing both carrier phase $\theta_0$ and amplitude $\rho_0$ (by definition $\rho_0$ is a positive real number).  Our aim is to estimate $a_0$ from $y_1, \dots, y_L$.  Complicating matters is that the transmitted symbols $s_i$ are not known to the reciever and must also be estimated.  Estimation problems of this type have undergone extensive prior study~\cite{ViterbiViterbi_phase_est_1983,Cowley_ref_sym_carr_1998,Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001}.  A practical approach is the least squares estimator, that is, the minimisers of the sum of squares function
\[
SS(a, \{s_i, i \in D\}) = \sum_{i = 1}^L \abs{ y_i - a s_i }^2 
\]
where $\abs{x}$ denotes the magnitude of the complex number $x$.  The least squares estimator is also the maximum likelihood estimator under the assumption that the noise sequence $\{w_i, i \in \ints\}$ is additive white and Gaussian.  As we will show, the estimator works well under less stringent assumptions.  Mackenthun~\cite{Mackenthun1994} described an algorithm to compute the least squares estimator that requires only $O(L \log L)$ arithmetic operations.  Sweldens~\cite{Sweldens2001} rediscovered Mackenthun's algorithm in 2001.

@@

\section{The least squares estimator}\label{sec:least-squar-estim}

As discussed, the least squares estimator is given by the minimiser of $SS(a, \{s_i, i \in D\})$ over all $a \in \complex$ and $s_1, \dots, s_L$ in the $M$-PSK constellation.  For fixed $a$ the least squares estimators of $s_1, \dots, s_L$, as functions $a = \rho e^{j \theta}$, are
\[
\hat{s}_i(a) = e^{j\hat{u}_i(\theta)} \qquad \text{where} \qquad \hat{u}_i(\theta) = \round{\angle( e^{-j\theta}y_i)},
\]
where $\angle(\cdot)$ denotes the complex argument (or phase), and $\round{\cdot}$ rounds its argument to the nearest multiple of $\frac{2\pi}{M}$.  Substituting $\hat{s}_i(a)$ into $SS(a, \{s_i, i \in D\})$, give the sum of squares, conditioned on minimisation with respect to the transmitted symbols,
\[
SS(a) = \sum_{i = 1}^L \abs{ y_i - a \hat{s}_i(a) }^2.
\]
The least squares estimator of the carrier phase $\hat{\theta}$ and amplitude $\hat{\rho}$ then satisfy $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ where $\hat{a} = \arg\min_{a \in \complex} SS(a)$.  Mackenthun~\cite{Mackenthun1994} described an algorithm that computes $\hat{a}$ using $O(L\log L)$ operations.

@@

\begin{theorem}\label{thm:normality} (Asymptotic normality)
Under the same conditions as Theorem~\ref{thm:consistency}, let $f(r,\phi)$ be the joint probability density function of $R_1$ and $\Phi_1$, and let
\[
g(\phi) = \int_{0}^{\infty} r f(r,\phi) dr.
\]
Put $\hat{\lambda}_L = \sfracpart{\theta_0 - \hat{\theta}}$ and $\hat{m}_L = \hat{\rho} - \rho_0 G(0)$. %If $g(\phi)$ is continuous at $\phi = \tfrac{2\pi}{M}k+\tfrac{\pi}{M}$ for each $k = 0, 1, \dots M-1$, then 
Then the distribution of $(\sqrt{L}\lambda_L, \sqrt{L}m_L)$ converges to the bivariate normal with zero mean and covariance
\[
\left( \begin{array}{cc} 
A H^{-2} & 0 \\
0 & B
\end{array} \right)
\]
as $L \rightarrow \infty$, where
\[
H = G(0) -  2 \sin(\tfrac{\pi}{M}) \sum_{k = 0}^{M-1} g(\tfrac{2\pi}{M}k + \tfrac{\pi}{M}),
\]
\[
A = \expect R_1^2\sin^2\fracpart{\Phi_1}, \;\;\; B = \expect R_1^2 \cos^2\fracpart{\Phi_1}. 
\]
\end{theorem}

\begin{corollary}
The mean square error of the amplitude estimator $\hat{\rho}$ is $\expect (\rho_0 - \hat{\rho})^2 = \frac{B}{L} + \rho_0^2(1 - G(0))^2$.
\end{corollary}

@@

We now make some discussion regarding the assumptions made by these theorems.

@@

Let the noise sequence $\{w_i\}$ be complex Gaussian with independent real and imaginary parts having zero mean and variance $\sigma^2$.  The joint pdf of the real and imaginary parts is then
\[
p(x,y) = \frac{1}{2\pi\sigma^2}e^{-(x^2 + y^2)/\sigma^2}
\]
Theorem~\ref{thm:consistency}~and~\ref{thm:normality} hold and, since the distribution of $w_1$ is circularly symetric, the distribution of $R_1e^{j\Phi_1}$ is identical to the distribution of $1 + \frac{1}{\rho_0} w_1$.
%and the joint density function of the real and imaginary parts of $Z$ is
%\[
%p_Z(x,y) = \rho_0^2 p( \rho_0 (x - 1), \rho_0 y ) = \frac{\kappa^2}{2\pi}e^{-\kappa^2(x ^2 - 2x + 1 + y^2) }
%\] 
%where $\kappa = \tfrac{\rho_0}{\sigma}$.  
It can be shown that
\[
g(\phi) = \frac{\cos^2(\phi)}{4\sqrt{\pi}}\left( \frac{e^{-\kappa^2} }{\sqrt{\pi}} + \kappa \Psi(\phi)  e^{-\kappa^2\sin^2(\phi)}\cos(\phi) \right)
\]
where
\[
\Psi(\phi) = 1 + \erf(\kappa \cos(\phi)) = 1 + \frac{2}{\sqrt{\pi}}\int_0^{\kappa \cos(\phi)} e^{-t^2} dt .
\]
The value of $A$ can be computed by numerical integration.

@@

\begin{figure}[p]
	\centering
		\includegraphics[width=\linewidth]{code/data/plotncM2-2.mps}
		\caption{Phase error for BPSK}
		\label{fig:plotphase}
\end{figure}

\begin{figure}[p]
	\centering
		\includegraphics[width=\linewidth]{code/data/plotncM4-2.mps}
		\caption{Phase error for QPSK}
		\label{fig:plotphase}
\end{figure}

\begin{figure}[p]
	\centering
		\includegraphics[width=\linewidth]{code/data/plotncM2-1.mps}
		\caption{Amplitude error for BPSK}
		\label{fig:plotphase}
\end{figure}

\begin{figure}[p]
	\centering
		\includegraphics[width=\linewidth]{code/data/plotncM4-1.mps}
		\caption{Amplitude error for QPSK}
		\label{fig:plotphase}
\end{figure}

>>>>

\begin{abstract}
We consider least squares estimators of carrier phase and amplitude from a noisy communications signal.  We focus on signaling constellations that have symbols evenly distributed on the complex unit circle, such as binary phase shift keying (BPSK), quaternary phase shift keying (QPSK), and $M$-PSK.  We show, under reasonably mild conditions on the distribution of the noise, that the least squares estimator of carrier phase is strongly consistent and asymptotically normaly distributed.  However, the amplitude estimator is not consistent, and we show that it converges to a positive real number that is a function of the true carrier amplitude, the noise distribution, and the size of the constellation.  The results of Monte Carlo simulations are provided and these corroborate the theoretical results.
\end{abstract}
\begin{IEEEkeywords}
Noncoherent detection, differential phase shift keying, asymptotic statistics
\end{IEEEkeywords}

@@

\newcommand{\calC}{\mathcal C}

In passband communications systems the transmitted signal typically undergoes time offset (delay), phase shift and attenuation.  These effects must be compensated for at the reciever. In this paper we assume that the time offset has been previously handled, and we focus on estimating the phase shift and attenuation.  We consider signalling constellations that have symbols evenly distributed on the complex unit circle, commonly refered to as $M$-ary phase shift keying ($M$-PSK).  In this case, the transmitted symbols take the form,
\[
s_i = e^{j u_i},
\]
where $j = \sqrt{-1}$ and $u_i$ is from the set $\{0, \tfrac{2\pi}{M}, \dots, \tfrac{2\pi(M-1)}{M}\}$ and $M$ is the size of the constellation.  %We denote the set of symbols contained in the $M$-PSK constellation by $\calC$.

We assume that time offset estimation and matched filtering have already been performed at the reciever.  The recieved symbols then take the form,
\begin{equation}\label{eq:sigmod}
y_i = a_0 s_i + w_i, \qquad i = 1, \dots, L,
\end{equation}
where $w_i$ is noise and $a_0 = \rho_0 e^{j\theta_0}$ is a complex number representing both carrier phase $\theta_0$ and amplitude $\rho_0$ (by definition $\rho_0$ is a positive real number).  Our aim is to estimate $a_0$ from $y_1, \dots, y_L$.  If the transmitted symbols $s_1, \dots, s_L$ are known apriory at the reciever then the least squares estimator is typically used,
\[
\hat{a}_{\text{uc}} = \arg \min_{a \in \complex} \sum_{i = 1}^L \abs{ y_i - a s_i }^2  = \frac{1}{L} \sum_{i = 1}^L y_i s_i^*,
\]
where $\complex$ is the set of complex numbers and $x^*$ and $\abs{x}$ denotes the complex conjugate and complex magnitude of the complex number $x$.  Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983} call this the \emph{unmodulated carrier} estimator.  This estimator can be used if the tranmitter includes \emph{pilot} symbols, known to the reciever, i.e. with~\emph{coherent detection}.

Here we are interested in \emph{noncoherent detection}, when $s_1, \dots, s_L$ are not known at the reciever, and must also be estimated.  Estimation problems of this type have undergone extensive prior study~\cite{ViterbiViterbi_phase_est_1983,Cowley_ref_sym_carr_1998,Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001,McKilliamLinearTimeBlockPSK2009}.  A practical approach is the least squares estimator,
\[
\hat{a} = \arg\min_{a \in \complex} \min_{s_1, \dots, s_L \in \calC} \sum_{i = 1}^L \abs{ y_i - a s_i }^2,
\]
where $\calC$ is the set of symbols from the $M$-PSK constellation.  The least squares estimator is also the maximum likelihood estimator under the assumption that the noise sequence $\{w_i, i \in \ints\}$ is additive white and Gaussian.  As we will show, the estimator works well under less stringent assumptions.  Mackenthun~\cite{Mackenthun1994} described an algorithm to compute the least squares estimator $\hat{a}$ that requires only $O(L \log L)$ arithmetic operations.  Sweldens~\cite{Sweldens2001} rediscovered Mackenthun's algorithm in 2001.

@@

% \section{The least squares estimator}\label{sec:least-squar-estim}

% As discussed, the least squares estimator is given by the minimiser of $SS(a, \{s_i, i \in D\})$ over all $a \in \complex$ and $s_1, \dots, s_L$ in the $M$-PSK constellation.  For fixed $a$ the least squares estimators of $s_1, \dots, s_L$, as functions $a = \rho e^{j \theta}$, are
% \[
% \hat{s}_i(a) = e^{j\hat{u}_i(\theta)} \qquad \text{where} \qquad \hat{u}_i(\theta) = \round{\angle( e^{-j\theta}y_i)},
% \]
% where $\angle(\cdot)$ denotes the complex argument (or phase), and $\round{\cdot}$ rounds its argument to the nearest multiple of $\frac{2\pi}{M}$.  Substituting $\hat{s}_i(a)$ into $SS(a, \{s_i, i \in D\})$, give the sum of squares, conditioned on minimisation with respect to the transmitted symbols,
% \[
% SS(a) = \sum_{i = 1}^L \abs{ y_i - a \hat{s}_i(a) }^2.
% \]
% The least squares estimator of the carrier phase $\hat{\theta}$ and amplitude $\hat{\rho}$ then satisfy $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ where $\hat{a} = \arg\min_{a \in \complex} SS(a)$.  Mackenthun~\cite{Mackenthun1994} described an algorithm that computes $\hat{a}$ using $O(L\log L)$ operations.

@@

\begin{theorem}\label{thm:normality} (Asymptotic normality)
Under the same conditions as Theorem~\ref{thm:consistency}, let $f(r,\phi)$ be the joint pdf of $R_1$ and $\Phi_1$, and let
\[
g(\phi) = \int_{0}^{\infty} r f(r,\phi) dr.
\]
Put $\hat{\lambda}_L = \sfracpart{\theta_0 - \hat{\theta}}$ and $\hat{m}_L = \hat{\rho} - \rho_0 G(0)$. %If $g(\phi)$ is continuous at $\phi = \tfrac{2\pi}{M}k+\tfrac{\pi}{M}$ for each $k = 0, 1, \dots M-1$, then 
Then the distribution of $(\sqrt{L}\hat{\lambda}_L, \sqrt{L}\hat{m}_L)$ converges to the bivariate normal with zero mean and covariance
\[
\left( \begin{array}{cc} 
H^{-2} A & 0 \\
0 & B
\end{array} \right)
\]
as $L \rightarrow \infty$, where
\[
H = G(0) -  2 \sin(\tfrac{\pi}{M}) \sum_{k = 0}^{M-1} g(\tfrac{2\pi}{M}k + \tfrac{\pi}{M}),
\]
\[
A = \expect R_1^2\sin^2\fracpart{\Phi_1}, \;\;\; B = \expect R_1^2 \cos^2\fracpart{\Phi_1}. 
\]
\end{theorem}

\begin{corollary}\label{cor:ampmse}
The mean square error of the amplitude estimator $\hat{\rho}$ is 
\[
\expect (\rho_0 - \hat{\rho})^2 = \frac{B}{L} + \rho_0^2(1 - G(0))^2 + o(1)
\] 
where $o(1)$ goes to zero as $L \rightarrow \infty$.
\end{corollary}
\begin{IEEEproof} Write
\begin{align*}
\expect (\rho_0 - \hat{\rho})^2 &= \expect \big( \rho_0(1 - G(0)) + G(0)\rho_0 - \hat{\rho} \big)^2 \\
&= \expect \big( \rho_0(1 - G(0)) + \hat{m}_L \big)^2 \\
&= \rho_0^2(1 - G(0))^2 + \expect \hat{m}_L^2 + 2\rho_0(1 - G(0))\expect\hat{m}_L,
\end{align*}
and the corollary holds since $\expect \hat{m}_L^2 = B/L + o(1)$ and $\expect\hat{m}_L = o(1)$ from Theorem~\ref{thm:normality}.
\end{IEEEproof}

@@

We now discuss the assumptions made by these theorems.  The assumption that the $w_1, \dots w_L$ are circularly symmetric can be relaxed, but this comes at the expense of making the theorem statement more complicated.  Primarily this is because if $w_i$ are not circularly symmetric then the distribution of the random variables $R_i$ and $\Phi_i$ may depend on $a_0$ and also on the transmitted symbols $s_1, \dots, s_L$.  The result of this is that the asymptotic variance described in Theorem~\ref{thm:normality} is more complicated, and depends on $a_0$ and $s_1, \dots, s_L$.  The circularly symmetric assumption may not always hold in practice, but it does lead to simple, and useful expressions for the asymptotic variance.

A key assumption in Theorem~\ref{thm:consistency} is that the function $\abs{G(x)}$ is uniquely maximised at $x = 0$.

@@

Let the noise sequence $\{w_i\}$ be complex Gaussian with independent real and imaginary parts having zero mean and variance $\sigma^2$.  The joint pdf of the real and imaginary parts is then
\[
p(x,y) = \frac{1}{2\pi\sigma^2}e^{-(x^2 + y^2)/\sigma^2}
\]
Theorem~\ref{thm:consistency}~and~\ref{thm:normality} hold and, since the distribution of $w_1$ is circularly symetric, the distribution of $R_1e^{j\Phi_1}$ is identical to the distribution of $1 + \frac{1}{\rho_0} w_1$.
%and the joint density function of the real and imaginary parts of $Z$ is
%\[
%p_Z(x,y) = \rho_0^2 p( \rho_0 (x - 1), \rho_0 y ) = \frac{\kappa^2}{2\pi}e^{-\kappa^2(x ^2 - 2x + 1 + y^2) }
%\] 
%where $\kappa = \tfrac{\rho_0}{\sigma}$.  
It can be shown that
\[
g(\phi) = \frac{\cos^2(\phi)}{4\sqrt{\pi}}\left( \frac{e^{-\kappa^2} }{\sqrt{\pi}} + \kappa \Psi(\phi)  e^{-\kappa^2\sin^2(\phi)}\cos(\phi) \right)
\]
where
\[
\Psi(\phi) = 1 + \erf(\kappa \cos(\phi)) = 1 + \frac{2}{\sqrt{\pi}}\int_0^{\kappa \cos(\phi)} e^{-t^2} dt .
\]
The value of $A$, $B$ and $H$ can be efficiently computed by numerical integration using the above formula for $g(\phi)$.

@@

We present the results of Monte-Carlo simulations that evaluate the performance of the least squares estimator.  In all simulations the noise $\{w_i\}$ is independent and identically distributed circularly symmetric and Gaussian with real and imaginary parts having variance $\sigma^2$.  Simulations are run with $M=2,4$ (BPSK and QPSK) and $L=16,256,4096$ with signal to noise ratio $\text{SNR} = \tfrac{\rho_0^2}{2\sigma^2}$ between \unit[-20]{dB} and \unit[20]{dB}.  Five thousand replications are used per experiment.

Figures~\ref{fig:plotphaseM2}~and~\ref{fig:plotphaseM4} show the mean square error (MSE) of the phase estimator $\hat{\theta}$.  The dots are the Monte-Carlo simulations of the least square estiamtor.  The dashed line is the simulated performance of least squares estimator when all the data symbols $s_1, \dots, s_L$ are known apriory at the reciever.  Viterbi calls this the \emph{unmodulated carrier} and we use this name also.  The dashed line acts as a lower bound on performance.  The solid line is the MSE predicted by Theorem~\ref{thm:normality}.  The Theorem accurately predicts the behaviour of the phase estimator when $L$ is large and when the SNR is not too small.  As the SNR decreases the variance of the phase estimator approaches that of the uniform distribution on $[-\pi/M, \pi/M)$, and Theorem~\ref{thm:normality} does not model this behaviour.

Figures~\ref{fig:plotphaseM2}~and~\ref{fig:plotphaseM4} show the mean square error (MSE) of the amplitude estimator $\hat{\rho}$.  The dots are again the Monte-Carlo simulations, and the dahsed line is the unmodulated carrier.  The solid line is the asymptotic MSE predicted by Corollary~\ref{cor:ampmse}.  The corollary accurately models the behaviour of the amplitude estimator when $L$ is sufficiently large.  The estimator is biased, and as $L\rightarrow\infty$, the MSE converges to $\rho_0^2(1 - G(0))^2$.  The effect is clearly seen in the Figures when the SNR is small.

%\section{Discussion}

\section{Conclusion}

\begin{figure}[p]
	\centering
		\includegraphics[width=\linewidth]{code/data/plotncM2-2.mps}
		\caption{Phase error for BPSK}
		\label{fig:plotphaseM2}
\end{figure}

\begin{figure}[p]
	\centering
		\includegraphics[width=\linewidth]{code/data/plotncM4-2.mps}
		\caption{Phase error for QPSK}
		\label{fig:plotphaseM4}
\end{figure}

\begin{figure}[p]
	\centering
		\includegraphics[width=\linewidth]{code/data/plotncM2-1.mps}
		\caption{Amplitude error for BPSK}
		\label{fig:plotampM2}
\end{figure}

\begin{figure}[p]
	\centering
		\includegraphics[width=\linewidth]{code/data/plotncM4-1.mps}
		\caption{Amplitude error for QPSK}
		\label{fig:plotampM}
\end{figure}
